---
title: "Inference formula reference"
subtitle: "One page of recognition cards for the estimation, testing, and Bayesian machinery used all term"
page_mode: reference
week_contract: reference
source_mode: source-informed
toc: true
---

## What this page is

This is a recognition card, not a new lesson. Each section below states one procedure's recipe in symbols,
then walks it once against the MAC Study's locked sample numbers so you can see the recipe actually run.
Nothing here is derived from scratch — the derivations live in the week notes (especially Weeks 4, 6, 9, and
12); this page exists so you can look a formula up mid-lab or mid-note without losing your place. Every
number below is carried over unchanged from the week that first established it, so if a number here and a
number in a week note ever look different, treat that as a flag to recheck, not two independent facts.

As the rest of the course does throughout, keep two kinds of numbers straight: the MAC Study's *sample data
in hand* (x̄ = 49.8, s = 15.2, n = 36 visit durations; p̂ = 0.38, n = 100 usage-survey respondents) is what
every recipe below actually plugs in. A *hypothetical "true" population value* (μ = 48, σ = 15, π = 0.35)
appears only where a recipe needs a stipulated truth to teach against — the Week 9 power calculation and the
Week 4 bias/variance demonstration — and is flagged as hypothetical each time it appears.

## Standard errors

The standard error of an estimator is the standard deviation of its sampling distribution — how much the
estimator would bounce around from sample to sample of the same size, drawn the same way.

| Estimator | Standard error formula | MAC Study value |
|---|---|---|
| Sample mean x̄ (σ known) | SE(x̄) = σ/√n | σ = 15, n = 36 → SE(x̄) = 15/6 = **2.5** |
| Sample proportion p̂ | SE(p̂) = sqrt( p̂(1 − p̂) / n ) | p̂ = 0.38, n = 100 → SE(p̂) = sqrt(0.38 × 0.62 / 100) = sqrt(0.002356) ≈ **0.0485** |

$$
\mathrm{SE}(\bar x) = \frac{\sigma}{\sqrt n}, \qquad
\mathrm{SE}(\hat p) = \sqrt{\frac{\hat p (1-\hat p)}{n}}
$$

Weeks 2 through 9 treat σ ≈ 15 as known for the visit-duration thread — a "known-σ" teaching simplification
standard at this level, stated explicitly whenever it is used. Week 10's bootstrap instead uses the sample
SD s = 15.2 directly (bootstrap SE ≈ s/√n ≈ 15.2/6 ≈ 2.53), a deliberately close-but-not-identical value to
the known-σ SE of 2.5 — a teaching point about estimated vs. known SE, not a discrepancy to resolve.

## Bias, variance, and MSE decomposition

Three ways to describe how an estimator misses the parameter it targets, and the identity linking them.

$$
\mathrm{Bias}(\hat\theta) = E[\hat\theta] - \theta, \qquad
\mathrm{MSE}(\hat\theta) = \mathrm{Var}(\hat\theta) + \mathrm{Bias}(\hat\theta)^2
$$

Week 4's derivation (full detail there) works two examples against the MAC Study, using the **hypothetical
true** μ = 48, σ = 15 as the stipulated truth a real analyst would never actually know:

- **The (n − 1) divisor.** The "natural" variance estimator that divides by n instead of (n − 1) has
  E[(1/n)Σ(xᵢ − x̄)²] = ((n−1)/n)σ² = (35/36)(225) = 218.75, so its bias is 218.75 − 225 = **−6.25**. This is
  the standard motivation for why the sample variance divides by (n − 1), not n.
- **A shrinkage estimator.** θ̂ = 0.9x̄, compared against the unbiased x̄, both targeting the hypothetical true
  μ = 48: Bias(θ̂) = 0.9(48) − 48 = **−4.8**; Var(θ̂) = 0.9² × Var(x̄) = 0.81 × 6.25 = **5.0625** (using
  Var(x̄) = SE(x̄)² = 2.5² = 6.25); MSE(θ̂) = 5.0625 + (−4.8)² = 5.0625 + 23.04 = **28.1025**. The unbiased
  x̄'s own MSE is just its variance, MSE(x̄) = Var(x̄) = **6.25** — far smaller. Here shrinkage is a clearly
  bad trade; the lesson is the decomposition, not that shrinkage is always bad.

## Likelihood and the MLE recipe

The likelihood L(θ) is the joint density/mass of the *observed, fixed* data, viewed as a function of the
unknown parameter θ. It is not a probability distribution over θ.

$$
L(\theta) = \prod_{i=1}^n f(x_i \mid \theta), \qquad
\ell(\theta) = \ln L(\theta), \qquad
\hat\theta_{\text{MLE}} = \arg\max_\theta \ell(\theta)
$$

**Recipe:** (1) write the joint density/mass of the data as a function of θ; (2) take the log to turn the
product into a sum; (3) differentiate with respect to θ and set equal to zero; (4) solve for θ̂ and confirm
it is a maximum.

Week 5 compares the Binomial likelihood kernel πᵏ(1−π)ⁿ⁻ᵏ at three candidate values of π, using the small
illustrative pilot batch (n = 5, k = 2 used the MAC — deliberately kept small for by-hand arithmetic, and
distinct from the full n = 100 survey):

| π | Kernel π²(1 − π)³ |
|---|---|
| 0.2 | 0.04 × 0.512 = **0.02048** |
| 0.4 | 0.16 × 0.216 = **0.03456** |
| 0.6 | 0.36 × 0.064 = **0.02304** |

π = 0.4 is highest of the three, foreshadowing the Week 6 result. Week 6 derives the MLE in closed form for
both families used all term:

- **Binomial:** π̂_MLE = k/n = 2/5 = **0.4**.
- **Normal, σ known:** μ̂_MLE = x̄, using a different small pilot batch of five visit durations (52, 46, 58,
  41, 53 minutes): (52+46+58+41+53)/5 = 250/5 = **50**.

## Confidence interval recipe

$$
\hat\theta \pm z^{*} \cdot \mathrm{SE}(\hat\theta)
$$

**Recipe:** (1) compute the point estimate; (2) compute its standard error; (3) pick a confidence level and
its z\* (z\* = 1.96 for 95%); (4) form estimate ± z\* × SE. Week 7 applies this to both threads:

- **CI for μ:** 49.8 ± 1.96(2.5) = 49.8 ± 4.9 → **(44.9, 54.7)**.
- **CI for π:** 0.38 ± 1.96(0.0485) = 0.38 ± 0.0951 → **(0.285, 0.475)**.

Remember the convention-risk flag from Week 7: a 95% CI's "95%" describes the long-run behavior of the
*procedure* across repeated samples, never "a 95% probability the parameter is in this particular interval."

## Test-statistic and p-value recipe

$$
z = \frac{\hat\theta - \theta_0}{\mathrm{SE}(\hat\theta)}, \qquad
p = P(\text{data this extreme or more} \mid H_0 \text{ true})
$$

**Recipe:** (1) state H₀ and Hₐ and pick α in advance; (2) compute the test statistic (estimate minus null
value, divided by SE); (3) convert the test statistic to a p-value using the sampling distribution under
H₀; (4) compare p to α and state the conclusion in terms of the data, never in terms of "P(H₀ true)."

Week 8 tests H₀: μ = 45 against Hₐ: μ ≠ 45, using last year's campus baseline (45 minutes) as the null value
against this term's sample (x̄ = 49.8, SE = 2.5):

$$
z = \frac{49.8 - 45}{2.5} = 1.92, \qquad p = 2\bigl(1 - \Phi(1.92)\bigr) \approx 2(0.0274) = \mathbf{0.0548}
$$

This fails to reject H₀ at α = 0.05 — a deliberately borderline result. As Week 8 stresses, this p-value is
P(data this extreme or more extreme | H₀ true), never P(H₀ true).

## Power recipe

$$
\text{Power} = 1 - \beta = P(\text{reject } H_0 \mid H_a \text{ true, for a specific alternative})
$$

**Recipe:** (1) find the critical value of the estimator under H₀ at the chosen α; (2) ask how likely the
estimator is to fall beyond that critical value if a *specific* alternative were true instead; (3) that
probability is the power against that alternative.

Week 9 computes power for a one-sided test of H₀: μ = 45 against the **hypothetical** alternative μ = 50, at
α = 0.05:

$$
\text{critical } \bar x = 45 + 1.645(2.5) = 49.11, \qquad
\text{Power} = P(\bar x > 49.11 \mid \mu = 50) = P(Z > -0.355) = \Phi(0.355) \approx \mathbf{0.639}
$$

so β ≈ 0.361. The alternative μ = 50 here is a stipulated hypothetical, used only to illustrate the
calculation — never claimed as a known fact about the MAC-visiting population.

## Bootstrap percentile-CI recipe

**Recipe:** (1) resample n observations **with replacement** from the observed sample, many times (thousands
of resamples); (2) compute the statistic of interest (e.g. the mean) on each resample, building up a
bootstrap distribution; (3) take that bootstrap distribution's standard deviation as an estimated SE, or take
its 2.5th and 97.5th percentiles directly as a percentile CI.

Week 10 applies this to the n = 36 visit-duration sample (x̄ = 49.8, s = 15.2), using the sample SD directly
rather than the known σ = 15:

$$
\mathrm{SE}_{\text{boot}} \approx \frac{s}{\sqrt n} = \frac{15.2}{6} \approx 2.53, \qquad
49.8 \pm 1.96(2.53) \approx \mathbf{(44.84,\ 54.76)}
$$

This interval sits close to, but is not identical to, the Week 7 CI of (44.9, 54.7) — the point of the
comparison is that the bootstrap's estimated SE (from s) tracks the known-σ SE (2.5) closely without
requiring σ to be known in advance.

## Permutation-test recipe

**Recipe:** (1) pool both groups' data together, ignoring group labels; (2) reshuffle the pooled data into
two groups of the original sizes, uniformly at random, and recompute the difference in group means; (3)
repeat many times to build a null distribution of differences under "no real group effect"; (4) the
permutation p-value is the fraction of shuffled differences at least as extreme as the one actually observed.

Week 11 applies this to the two-group workshop comparison: workshop group n₁ = 20 (mean = 52.4 min,
SD = 10 min) vs. control group n₂ = 20 (mean = 45.1 min, SD = 10 min), observed difference = **7.3 minutes**.
A normal-approximation cross-check gives

$$
\mathrm{SE}_{\text{diff}} = \sqrt{\frac{10^2}{20} + \frac{10^2}{20}} = \sqrt{10} \approx 3.162, \qquad
z = \frac{7.3}{3.162} \approx 2.31, \qquad p \approx \mathbf{0.021}
$$

Week 11's simulated permutation p-value is reported as closely matching this approximation (≈ 0.02), as
expected when the two groups' sizes and spreads are this close to equal.

## Beta-Binomial Bayesian update recipe

$$
\pi(\theta) \sim \mathrm{Beta}(a, b) \ \longrightarrow\ \pi(\theta \mid \text{data}) \sim
\mathrm{Beta}(a + k,\ b + n - k)
$$

**Recipe:** (1) state a Beta(a, b) prior on the proportion π; (2) observe k successes in n trials; (3) the
posterior is exactly Beta(a + k, b + n − k) — the Beta family is conjugate to the Binomial, so no numerical
integration is needed; (4) summarize the posterior with its mean, variance, or a credible interval.

Week 12 updates a Beta(a = 3, b = 7) prior (prior mean 3/10 = 0.30, representing the MAC director's belief
before this term's survey) using the full survey's n = 100, k = 38:

$$
\mathrm{Beta}(3, 7) \ \longrightarrow\ \mathrm{Beta}(3+38,\ 7+62) = \mathrm{Beta}(41, 69)
$$

Posterior mean = 41/110 ≈ **0.373**. Posterior variance = (41 × 69) / (110² × 111) ≈ 0.002106, so posterior
SD ≈ **0.0459**. A normal-approximation 95% credible interval is

$$
0.373 \pm 1.96(0.0459) \approx \mathbf{(0.283,\ 0.463)}
$$

Keep π(θ) (the Bayesian prior/posterior density) visually and verbally distinct from π (the population
proportion itself) and from L(θ) (the likelihood, a function of θ but not a density over θ) — context
disambiguates, but the distinction is worth restating every time all three appear together.

## In R

The chunks below are shown as teaching reference only — they are not executed in this build (`#| eval:
false`), are base-R only, and use `set.seed(35103)` wherever a simulation is involved, so they are
reproducible when a student runs them in their own session. The R distribution-function family follows a
consistent prefix + root naming pattern: `d`-root gives a density/mass (height of the curve), `p`-root gives
a cumulative probability (area to the left), `q`-root gives a quantile (the x-value for a given cumulative
probability — the inverse of `p`-root), and `r`-root generates random draws. The root itself names the
distribution (`norm`, `binom`, `beta`, and so on).

```{r}
#| label: se-and-ci
#| eval: false
set.seed(35103)

# Standard errors
se_xbar <- 15 / sqrt(36)                 # SE(x-bar), known sigma = 15
p_hat   <- 0.38
se_phat <- sqrt(p_hat * (1 - p_hat) / 100)

# z* for a 95% CI comes from the standard normal quantile function
z_star <- qnorm(0.975)                   # q-root: quantile for a given cumulative probability

ci_mu <- 49.8 + c(-1, 1) * z_star * se_xbar
ci_pi <- p_hat + c(-1, 1) * z_star * se_phat
```

```{r}
#| label: p-value-and-power
#| eval: false
set.seed(35103)

# p-value for z = 1.92, two-sided
z_stat  <- (49.8 - 45) / 2.5
p_value <- 2 * (1 - pnorm(z_stat))        # p-root: cumulative probability up to z_stat

# Power against the hypothetical alternative mu = 50
crit_xbar <- 45 + qnorm(0.95) * 2.5
power     <- 1 - pnorm(crit_xbar, mean = 50, sd = 2.5)
```

```{r}
#| label: likelihood-kernel
#| eval: false
set.seed(35103)

# Binomial likelihood kernel at three candidate values of pi, n = 5, k = 2
pis     <- c(0.2, 0.4, 0.6)
kernels <- dbinom(2, size = 5, prob = pis)  # d-root: mass/density at the observed data

# MLE by direct search over a fine grid (Week 6 derives the closed form pi-hat = k/n)
grid       <- seq(0.001, 0.999, by = 0.001)
loglik     <- dbinom(2, size = 5, prob = grid, log = TRUE)
mle_pi_hat <- grid[which.max(loglik)]
```

```{r}
#| label: bootstrap-percentile-ci
#| eval: false
set.seed(35103)

# Illustrative: bootstrap resampling from a stand-in sample with x-bar = 49.8, s = 15.2, n = 36
visit_sample <- rnorm(36, mean = 49.8, sd = 15.2)  # a stand-in draw; the course's actual n=36 sample is fixed

boot_means <- replicate(2000, {
  resample <- sample(visit_sample, size = 36, replace = TRUE)  # resampling WITH replacement
  mean(resample)
})

boot_ci <- quantile(boot_means, probs = c(0.025, 0.975))  # the percentile CI
```

```{r}
#| label: permutation-test
#| eval: false
set.seed(35103)

# Illustrative: permutation test for the workshop vs. control comparison
workshop <- rnorm(20, mean = 52.4, sd = 10)
control  <- rnorm(20, mean = 45.1, sd = 10)
observed_diff <- mean(workshop) - mean(control)

pooled <- c(workshop, control)
n1     <- length(workshop)

perm_diffs <- replicate(5000, {
  shuffled <- sample(pooled)              # reshuffle the pooled data uniformly at random
  mean(shuffled[1:n1]) - mean(shuffled[(n1 + 1):length(pooled)])
})

perm_p_value <- mean(abs(perm_diffs) >= abs(observed_diff))
```

```{r}
#| label: beta-binomial-update
#| eval: false
set.seed(35103)

# Beta(3, 7) prior updated by n = 100, k = 38
a_prior <- 3; b_prior <- 7
k <- 38; n <- 100

a_post <- a_prior + k
b_post <- b_prior + (n - k)

post_mean <- a_post / (a_post + b_post)
post_draws <- rbeta(10000, a_post, b_post)  # r-root: random draws from the posterior

post_ci <- quantile(post_draws, probs = c(0.025, 0.975))  # simulation-based credible interval
```

## Public vs. graded

These notes, the examples, and the practice here are **public and ungraded** — study material only. No
graded prompts, answer keys, rubrics, point values, or due dates appear on this site. Graded inference
checkpoints, quizzes, homework, labs, the midterm, the project, and the final live in **Blackboard
(the LMS)**, which is authoritative for due dates, submissions, and grades. If this page and Blackboard
ever disagree, follow Blackboard.

## See also

- [Notation glossary](notation-glossary.qmd) — every symbol used on this page (θ, θ̂, SE, L(θ), π(θ), and the
  rest) defined in one place.
- [R and Quarto setup](r-quarto-setup.qmd) — how to get the course's R and Quarto environment working before
  running any of the chunks above yourself.
- [Week 4 — Bias, variance, and mean squared error](../notes/week-04-bias-variance-and-mean-squared-error.qmd)
- [Week 5 — Likelihood](../notes/week-05-likelihood.qmd)
- [Week 6 — Maximum likelihood estimation](../notes/week-06-maximum-likelihood-estimation.qmd)
- [Week 7 — Confidence intervals (+ midterm)](../notes/week-07-confidence-intervals-and-midterm.qmd)
- [Week 8 — Hypothesis tests and p-values](../notes/week-08-hypothesis-tests-and-p-values.qmd)
- [Week 9 — Error rates, power, and decisions](../notes/week-09-error-rates-power-and-decisions.qmd)
- [Week 10 — Bootstrap inference](../notes/week-10-bootstrap-inference.qmd)
- [Week 11 — Randomization and permutation tests](../notes/week-11-randomization-and-permutation-tests.qmd)
- [Week 12 — Bayesian inference](../notes/week-12-bayesian-inference.qmd)
