---
title: "Lab 10 — Bootstrap intervals"
subtitle: "Building a percentile confidence interval by resampling, by hand, in base R"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab is the hands-on companion to
> [Week 10 — Bootstrap inference](../notes/week-10-bootstrap-inference.qmd). The week note develops the
> percentile bootstrap conceptually and compares it to the Week 7 theoretical confidence interval; this lab
> carries that same idea out step by step in code, so you can watch a bootstrap distribution build itself one
> resample at a time.

## The idea

A confidence interval built from a formula, like the Week 7 interval for the MAC Study's mean visit duration,
leans on a theoretical result: if you know (or are willing to assume) the shape of the sampling distribution
of $\bar{x}$, you can write down its standard error directly and use a normal quantile to build the interval.
The bootstrap takes a different route. It says: you do not actually have the population, and you do not want
to lean on a normal-theory formula for the standard error of some more complicated statistic — so instead,
**treat the sample itself as a stand-in for the population**, and repeatedly resample from it, with
replacement, to see how much a statistic computed from a sample of this size actually bounces around.

Concretely: draw a new sample of size $n$ from your original $n$ observations, sampling **with replacement**
(so the same original observation can appear zero, one, or several times in a given resample). Compute the
statistic of interest — here, the mean — on that resample. Repeat this thousands of times. The resulting
collection of resample means is the **bootstrap distribution** of $\bar{x}$. Its spread approximates the true
sampling distribution's spread, and its middle 95% (the 2.5th to 97.5th percentile) gives a **percentile
bootstrap confidence interval** — no normal-theory formula required.

This lab uses base R only, following this course's house convention (`sample()`, `replicate()`, and friends)
rather than the `infer`-package pipeline that ModernDive's own examples use. The framing — build the sampling
distribution of a statistic by simulation rather than by formula — is exactly ModernDive's simulation-based
approach; we are just writing it out in base R instead of `specify() |> generate() |> calculate()`.

## Goal

By the end of this lab you will have:

1. Constructed a synthetic stand-in sample for the MAC Study's $n = 36$ visit-duration sample.
2. Drawn $B = 2000$ bootstrap resamples from it, with replacement, and computed the mean of each.
3. Read off the 2.5th and 97.5th percentiles of the bootstrap distribution of means as a 95% percentile
   confidence interval.
4. Compared that bootstrap interval to the Week 7 theoretical confidence interval, $(44.9, 54.7)$.

## Setup

All chunks below are shown with `#| eval: false` — this build presents code as teaching material, not as
executed output. When you run this in your own R or Posit Cloud session, keep `set.seed(35103)` at the top so
your results are reproducible and comparable to a classmate's.

```{r}
#| label: setup
#| eval: false
set.seed(35103)

# No external packages needed — base R only, per course convention.
```

A note on the data before you start: the MAC Study's real recorded sample is $n = 36$ visits with sample mean
$\bar{x} = 49.8$ minutes and sample SD $s = 15.2$ minutes (Week 7, Week 10). This lab does not have access to
the 36 individual recorded visit lengths — only their summary. So Step 1 below builds a **synthetic
stand-in** vector: 36 numbers, drawn from a normal model, engineered to have (very nearly) the same mean and
SD as the real sample. This stand-in is clearly a teaching device, not the real data — but it behaves the
same way under resampling, which is all this lab needs.

## Steps

### Step 1 — Build a synthetic stand-in sample

Generate 36 synthetic values from a Normal(49.8, 15.2) model. This is **not** the real recorded sample of 36
MAC visits — it is a synthetic vector engineered to share the real sample's mean and SD, standing in for it so
the resampling mechanics below have something concrete to work on.

```{r}
#| label: synthetic-sample
#| eval: false
set.seed(35103)

# Synthetic stand-in for the real n = 36 visit-duration sample.
# Engineered to share the real sample's summary: mean 49.8, SD 15.2.
visit_sample <- rnorm(36, mean = 49.8, sd = 15.2)

length(visit_sample)   # should be 36
mean(visit_sample)     # near 49.8, not exactly (rnorm() draws are random)
sd(visit_sample)       # near 15.2, not exactly
```

**This chunk is shown but not executed on this page** (`eval: false`) — running it yourself is the point. The
figure below was produced by running exactly this chunk's code separately (same seed `35103`, same
`rnorm(36, mean = 49.8, sd = 15.2)` call), so you can see what `visit_sample` actually looks like as 36 points
rather than only as two summary numbers:

![**The synthetic stand-in sample, visualized (synthetic).** 36 points drawn from Normal(49.8, 15.2), the
stand-in this lab treats as "the sample in hand." The dashed line marks this particular draw's mean
(49.09), close to but not exactly the 49.8
target.](../assets/figures/lab10_synthetic_sample.png){#fig-lab10-synthetic-sample fig-alt="A one-dimensional dot plot (jittered) of 36 points scattered mostly between about 25 and 80 minutes, with a dashed vertical line marking the sample mean near 49."}

Because `rnorm()` draws are random, `mean(visit_sample)` and `sd(visit_sample)` will land close to 49.8 and
15.2 but not hit them exactly — that is expected and fine. What matters for the rest of this lab is that
`visit_sample` is a length-36 numeric vector in the right neighborhood, which we now treat as "the sample in
hand" for resampling purposes.

### Step 2 — Draw one bootstrap resample

Before automating thousands of resamples, look at a single one. Sample 36 values from `visit_sample`, **with
replacement**, and compute the mean of that resample.

```{r}
#| label: one-resample
#| eval: false
set.seed(35103)

one_resample <- sample(visit_sample, size = 36, replace = TRUE)

length(one_resample)     # still 36 — same size as the original
mean(one_resample)       # a single bootstrap replicate of the mean
```

**This chunk is shown but not executed on this page** — running it yourself is the point. The figure below was
produced by running exactly this chunk's code separately (same seed `35103`), then counting how many times
each of the 36 original values landed in `one_resample`, so you can see "with replacement" as a picture rather
than only as a phrase:

![**Sampling with replacement, made visible (synthetic).** For this one resample, 11 of the 36 originals were
skipped entirely (drawn 0 times), 17 were drawn exactly once, and 8 were drawn two or more times — exactly the
mechanism `replace = TRUE`
produces.](../assets/figures/lab10_one_resample.png){#fig-lab10-one-resample fig-alt="A bar chart of 36 bars, one per original observation sorted low to high, showing how many times each was drawn in one resample. Most bars are height 0 or 1; a handful reach height 2 or 3, colored darker to mark them as drawn multiple times."}

Notice `replace = TRUE` is doing the essential work here: without it, `sample()` would just reshuffle the same
36 values in a different order and every resample mean would come out identical to `mean(visit_sample)`. With
replacement, some original values are skipped in a given resample and others are drawn two, three, or more
times, so the resample mean genuinely varies from resample to resample. That variation is exactly what the
bootstrap is measuring.

### Step 3 — Automate B = 2000 bootstrap resamples

Use `replicate()` to repeat Step 2's logic 2000 times, storing each resample's mean.

```{r}
#| label: bootstrap-replicate
#| eval: false
set.seed(35103)

B <- 2000

boot_means <- replicate(
  B,
  mean(sample(visit_sample, size = 36, replace = TRUE))
)

length(boot_means)   # should be 2000
head(boot_means)      # first few bootstrap replicate means
mean(boot_means)      # center of the bootstrap distribution
sd(boot_means)        # bootstrap standard error of the mean
```

`boot_means` is now a vector of 2000 numbers, each one the mean of an independent resample-with-replacement of
size 36 drawn from `visit_sample`. This vector *is* the bootstrap distribution. Its standard deviation,
`sd(boot_means)`, is the **bootstrap standard error** — an estimate of how much $\bar{x}$ would vary from
sample to sample, built entirely by resampling rather than by a formula. It should land in the neighborhood of
$s/\sqrt{n} = 15.2/6 \approx 2.53$, the bootstrap SE reported in the Week 10 note, though the exact value will
depend on the particular `visit_sample` draw from Step 1.

**This chunk is shown but not executed on this page.** The figure below extends this chunk's already-computed
`boot_means` with a plotting call not written on the page — tracking the *cumulative* mean and SD as
resamples accumulate, produced by running exactly this chunk's code separately (same seed, same `B = 2000`),
so you can watch both settle down toward a stable value rather than only reading off their final numbers:

![**Bootstrap replication converging (synthetic).** As the number of accumulated resamples grows from 1 to
2000, the running mean of `boot_means` settles near the original sample mean (left), and the running SD
settles near the theoretical standard error $s/\sqrt{n} \approx 2.34$ (right) — this settling-down *is* what
"repeat this thousands of times" is
buying.](../assets/figures/lab10_boot_convergence.png){#fig-lab10-boot-convergence fig-alt="Two side-by-side line charts. The left chart shows the cumulative mean of the bootstrap resample means leveling off near the original sample mean as more resamples accumulate. The right chart shows the cumulative standard deviation of the same resample means leveling off near the theoretical standard error."}

**See the distribution, not just its summary.** A histogram of `boot_means` is the actual pedagogical payoff
of this lab: watching 2000 resample means pile up into a distribution, rather than only reading off two
summary numbers.

```{r}
#| label: bootstrap-distribution-histogram
#| eval: false

hist(boot_means, breaks = 40,
     main = "Bootstrap distribution of the mean (B = 2000 resamples)",
     xlab = "resample mean visit duration (minutes)")
abline(v = mean(visit_sample), lty = 2)   # centered near the original sample mean, ~49.8
```

**What this shows (public-safe description, since this chunk is not executed here):** a roughly bell-shaped
histogram of the 2000 resample means, centered near the original sample mean ($\approx 49.8$, marked by the
dashed line) and spread out by roughly `sd(boot_means)` in either direction — this shape *is* the bootstrap
distribution Step 4 reads its percentile interval from, made visible rather than only summarized by two
numbers. The figure below was produced by running exactly this chunk's `hist()`/`abline()` code separately
(same seed, same `boot_means`), so you can check your prediction of its shape against a real result:

![**The bootstrap distribution, rendered (synthetic).** A histogram of the 2000 bootstrap resample means, with
a dashed line at the original sample mean (49.09 for this particular `visit_sample` draw). The bell shape and
the spread are exactly what Step 4 reads its percentile interval
from.](../assets/figures/lab10_boot_hist.png){#fig-lab10-boot-hist fig-alt="A bell-shaped histogram of 2000 bootstrap resample means, roughly symmetric, with a dashed vertical line near the center marking the original sample mean."}

### Step 4 — Read off the percentile confidence interval

The percentile bootstrap 95% confidence interval is simply the 2.5th and 97.5th percentiles of `boot_means`.

```{r}
#| label: percentile-ci
#| eval: false
set.seed(35103)

boot_ci <- quantile(boot_means, probs = c(0.025, 0.975))

boot_ci
```

`quantile(boot_means, probs = c(0.025, 0.975))` cuts off the lowest 2.5% and the highest 2.5% of the 2000
bootstrap means, leaving the middle 95% — read directly off the simulated distribution, with no normal-theory
formula anywhere in sight. Compare the two numbers `boot_ci` prints to the Week 10 note's reported percentile
bootstrap interval, approximately $(44.84, 54.76)$, and to the Week 7 theoretical interval, $(44.9, 54.7)$.
They should be close but not identical: the bootstrap interval is built from resampling a single synthetic
stand-in sample, so its exact endpoints depend on that sample's particular values and on which 2000 resamples
happened to be drawn, while the Week 7 interval comes from a normal-theory formula using the known-$\sigma$
simplification. Rerunning Steps 1–4 with a different seed would shift `boot_ci` slightly; the overall center
and width should stay recognizably similar.

**This chunk is shown but not executed on this page.** The figure below extends this chunk's `boot_ci` object
with a comparison plot not written on the page, produced by running exactly this chunk's code separately (same
seed, same `boot_means`):

![**The percentile bootstrap CI vs. the Week 7 theoretical CI (synthetic).** For this particular run,
`boot_ci` = (44.62, 53.58) — close to, but not identical to, the Week 7 theoretical interval (44.9, 54.7);
both are reported here rather than treating one as simply replacing the
other.](../assets/figures/lab10_ci_comparison.png){#fig-lab10-ci-comparison fig-alt="A horizontal number line from about 41 to 58 minutes with two bracket intervals stacked vertically: the Week 7 theoretical 95 percent CI on top, and the Step 4 percentile bootstrap 95 percent CI below it, overlapping substantially."}

## Verify

Before treating your bootstrap interval as trustworthy, check all of the following in your own session:

- `length(visit_sample)` is exactly 36, and `mean(visit_sample)` / `sd(visit_sample)` land close to 49.8 and
  15.2 respectively (not exact, but in the right neighborhood — large discrepancies mean Step 1 was not run
  correctly, or `set.seed(35103)` was skipped or placed inconsistently).
- `length(boot_means)` is exactly 2000 — one mean per bootstrap resample, no more, no fewer.
- `boot_ci` is a two-number interval with the lower bound below the upper bound, both numbers in the general
  neighborhood of the 40s-to-50s range (not, for instance, negative or in the hundreds — a sign something in
  Step 1 or Step 3 was mis-specified).
- The width of `boot_ci` (upper minus lower) is in the same rough neighborhood as the Week 7 interval's width
  of $54.7 - 44.9 = 9.8$ minutes — the bootstrap and theoretical intervals should agree approximately, not
  wildly diverge, because both are estimating the same quantity (the sampling variability of $\bar{x}$ for
  $n = 36$) from consistent inputs.
- Re-running the whole script from `set.seed(35103)` at the top reproduces the same `boot_ci` numbers exactly
  — if it does not, some randomness-consuming step is happening before the seed is set, or out of order.

Consider a small, unrelated synthetic scenario, useful for sanity-checking the resampling logic without
re-running code: a teaching lab (not the MAC Study) records 5 quiz-completion times, in minutes:
$8, 9, 11, 12, 15$ — synthetic numbers, not drawn from any real class. Their mean is
$(8+9+11+12+15)/5 = 55/5 = 11$. One possible bootstrap resample of size 5, drawn with replacement, might be
$9, 11, 11, 8, 15$ (the value 11 happened to be drawn twice, and 12 was skipped this time); its mean is
$(9+11+11+8+15)/5 = 54/5 = 10.8$. Another resample might draw $8, 8, 12, 15, 15$, with mean
$(8+8+12+15+15)/5 = 58/5 = 11.6$. Repeating this thousands of times and collecting the resample means builds
a bootstrap distribution centered near 11, with spread reflecting how much a mean of 5 values drawn from this
small pool can plausibly vary — exactly the same mechanism as Steps 2–4 above, just small enough to trace by
hand for a couple of resamples. This synthetic 5-value example is unrelated to the MAC Study's likelihood
pilot batch from Weeks 5–6, which also happens to use 5 observations — the coincidence in size is not
meaningful; the values and context here are freshly invented for this lab.

## AI use note

| Tool | Purpose | Verification |
|------|---------|---------------|
| None used for this lab's code or numbers | All base-R chunks, the synthetic stand-in sample, and the hand-traced transfer example were authored directly against this course's own conventions | Every locked figure (the Week 7 interval $(44.9, 54.7)$, the Week 10 bootstrap interval $\approx(44.84, 54.76)$, and the bootstrap SE $\approx 2.53$) was checked against the course's own recurring-case numbers rather than re-derived; the transfer-example arithmetic was checked by hand ($55/5=11$; $54/5=10.8$; $58/5=11.6$) |

If you use an AI tool to help you debug or extend this lab in your own work, record what you asked it, what it
returned, and how you checked the result before trusting it — the same two-column habit modeled above.

## See also

- [Week 10 — Bootstrap inference](../notes/week-10-bootstrap-inference.qmd) — the concept companion to this
  lab.
- [Inference formula reference](../resources/inference-formula-reference.qmd)
- [R & Quarto setup](../resources/r-quarto-setup.qmd)

The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.
