---
title: "Lab 2 — Simulating sampling distributions"
subtitle: "Watching x̄ ~ Normal(48, 2.5²) happen, instead of just writing it down"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** Week 2's note (see
> [Week 2 — Sampling distributions and simulation](../notes/week-02-sampling-distributions-and-simulation.qmd))
> derives that, under the hypothetical true world Normal(μ = 48, σ = 15), the sampling distribution of the
> mean of n = 36 MAC visit durations is approximately Normal(48, 2.5²), where 2.5 = σ/√n = 15/√36 = 15/6. That
> claim is a statement about *infinitely many* hypothetical samples you will never actually collect. This lab
> makes it concrete by simulating a large but finite number of samples yourself, computing the mean of each
> one, and looking at what the resulting collection of means actually looks like — its center, its spread, and
> its shape — next to the theoretical curve the note derived. You are not proving the claim here; you are
> watching it happen.

## The idea

A sampling distribution is the distribution of a statistic (here, x̄) across *repeated* samples from the same
world. You can never actually take infinitely many repeated samples of real MAC visits — you get one sample,
once. But a computer can pretend to be the hypothetical true world as many times as you like, because
Normal(μ = 48, σ = 15) is a world you get to fully specify yourself. This lab uses that trick: it draws many
independent samples of size n = 36 from Normal(48, 15), computes the mean of each simulated sample, and then
looks at the distribution of those means. If the Week 2 derivation is right, this simulated distribution of
means should end up centered near 48 with a standard deviation near 2.5, and its shape should look Normal. The
sample data in hand from the real MAC survey (x̄ = 49.8, n = 36) is never used inside the simulation itself —
the simulation only ever draws from the *hypothetical* true world, exactly as flagged in the note. Comparing
the simulated distribution to the theoretical Normal(48, 2.5²) curve is the whole point: it turns an algebraic
formula (σ/√n) into something you can see settle into place as the number of simulated samples grows.

This is also a place where the course's simulation-based framing (informed by ModernDive's emphasis on seeing
a sampling distribution by resampling) meets the course's base-R-only convention (§5 of the build
conventions): the idea of "generate many samples, compute a statistic from each, look at the resulting
distribution" is exactly ModernDive's simulation-based inference framing, but every step below is shown with
base R (`rnorm()`, a loop or `replicate()`, `mean()`) rather than the `infer` package's `specify()` /
`generate()` / `calculate()` pipeline.

![**This lab's simulation pipeline (synthetic).** Five steps, always in this order: start from a fully
specified hypothetical world, draw many independent samples from it, collapse each sample to its own mean,
look at the resulting collection of means as a histogram, and compare that histogram to the theoretical curve
the Week 2 derivation predicts.](../assets/figures/lab02_sim_pipeline.png){#fig-lab02-pipeline fig-alt="A left-to-right flow diagram of five boxes connected by arrows: hypothetical world Normal 48 15, draw B=5000 samples of n=36, colMeans to get 5000 x-bar values, histogram of the 5000 means, compare to Normal 48 2.5 squared."}

## Goal

By the end of this lab you should be able to:

- Write base-R code that draws `B` independent random samples of size `n` from a specified Normal world.
- Compute the sample mean of each simulated sample and collect the results into a single vector of length `B`.
- Summarize that vector (mean, SD) and compare the summary numbers to the theoretical mean and standard error
  from the Week 2 derivation.
- Plot a histogram of the simulated means and overlay the theoretical Normal density, and describe in words
  how the two compare.

## Setup

You will need base R only — no packages beyond what ships with R. Before you run anything below in your own
session (remember: in this build the chunks are shown, not executed, so you will run them yourself later),
set your seed so your results match the ones described here:

```{r}
#| label: setup
#| eval: false
set.seed(35103)
```

Keep this `set.seed(35103)` call at the top of your script, before you draw any random numbers. If you re-run
the whole script from the top with the same seed, you will get the same simulated numbers every time — that
is what "reproducible" means in a simulation context, and it is why every simulation chunk in this course
carries the same seed.

Recall from the Week 2 note the two numbers that anchor this whole lab, both hypothetical-world quantities,
never the real sample facts:

- Hypothetical true world: Normal(μ = 48, σ = 15) — a stipulated truth used only for this teaching device.
- Sample size: n = 36 (matching the size of the real MAC visit-duration sample, so the comparison to that
  sample's standard error is apples to apples).
- Theoretical sampling distribution of x̄, from the Week 2 derivation: Normal(48, 2.5²), since
  SE(x̄) = σ/√n = 15/√36 = 15/6 = 2.5.

## Steps

### Step 1 — Simulate B = 5000 samples of size 36 from Normal(48, 15)

The core idea is to draw not one sample of 36 values, but many independent samples of 36 values each, all from
the same hypothetical true world. A convenient way to organize this in base R is to draw one long vector of
`B * n` values, then arrange it into a matrix with `n` rows and `B` columns, so that each *column* is one
simulated sample of size 36.

```{r}
#| label: simulate-samples
#| eval: false
set.seed(35103)

mu_true    <- 48
sigma_true <- 15
n          <- 36
B          <- 5000

# One long draw of B * n independent Normal(48, 15) values, reshaped so each
# column is one simulated sample of size n = 36.
sim_values  <- rnorm(B * n, mean = mu_true, sd = sigma_true)
sim_samples <- matrix(sim_values, nrow = n, ncol = B)
```

After this step, `sim_samples` is a 36-by-5000 matrix. Each of the 5000 columns is a separate, independent
pretend sample of 36 MAC visit durations, drawn from the hypothetical Normal(48, 15) world — not from the real
survey data, and not from each other.

### Step 2 — Compute the sample mean of each simulated sample

Now collapse each column of `sim_samples` down to a single number: that column's sample mean. Base R's
`colMeans()` does exactly this in one line, giving you a vector of 5000 simulated sample means — one x̄ value
per pretend sample.

```{r}
#| label: compute-means
#| eval: false
set.seed(35103)

xbar_sim <- colMeans(sim_samples)

length(xbar_sim)   # should be 5000
head(xbar_sim)      # first few simulated sample means
```

`xbar_sim` is now the object of interest for the rest of the lab: it is a simulated approximation to the
sampling distribution of x̄. Each of its 5000 entries is what you would have gotten as your sample mean, had
that particular pretend sample of 36 visits been the one you actually drew from the hypothetical true world.

### Step 3 — Summarize the distribution and plot a histogram

With 5000 simulated means in hand, compute their mean and standard deviation, and compare both numbers to the
theoretical values the Week 2 note derived: a mean at 48 and a standard error of 2.5.

```{r}
#| label: summarize-and-plot
#| eval: false
set.seed(35103)

mean(xbar_sim)   # expect a number close to 48
sd(xbar_sim)     # expect a number close to 2.5

hist(xbar_sim,
     breaks = 40,
     freq   = FALSE,
     main   = "Simulated sampling distribution of x-bar (B = 5000, n = 36)",
     xlab   = "simulated sample mean")
```

**This chunk is shown but not executed on this page** (`eval: false`) — running it yourself in your own R
session, with `set.seed(35103)` restored at the top, is the point of this lab. The figure below was produced
by running exactly this simulation separately (same seed, same `mu_true`/`sigma_true`/`n`/`B`), so you can
compare its shape and its printed `mean(xbar_sim)`/`sd(xbar_sim)` values to what you get when you run it
yourself:

![**The simulated sampling distribution, before the theoretical curve (synthetic).** A histogram of 5000
simulated sample means from Normal(48, 15) with n = 36, on a density scale. The dashed line marks the
simulation's own mean, 47.973 — close to, but not exactly, the theoretical 48; `sd(xbar_sim)` came out to
2.520, close to the theoretical SE of
2.5.](../assets/figures/lab02_sim_hist.png){#fig-lab02-sim-hist fig-alt="A bell-shaped histogram of 5000 simulated sample means, centered near 48, with a dashed vertical line marking the simulation's own mean near 48 and text reporting the simulated mean and standard deviation."}

Because `B = 5000` is large but finite, `mean(xbar_sim)` will not land on exactly 48, and `sd(xbar_sim)` will
not land on exactly 2.5 — expect small simulation noise, typically numbers such as 47.9-something and
2.4-something-to-2.5-something, changing slightly if you rerun with a different seed or a different `B`. The
point of the exercise is that these simulated numbers land *close to* the theoretical mean and SE, not that
they match them digit for digit. The histogram itself, with `freq = FALSE` so it is on a density scale, should
already look bell-shaped and centered near 48 even before you add the theoretical curve in Step 4.

### Step 4 — Overlay the theoretical Normal(48, 2.5²) curve

Finally, draw the theoretical density curve for Normal(48, 2.5²) on top of the histogram, using base R's
`dnorm()` and `curve()`, so you can see directly how closely the simulated distribution tracks the formula
from the Week 2 derivation.

```{r}
#| label: overlay-theoretical-curve
#| eval: false
set.seed(35103)

se_theoretical <- sigma_true / sqrt(n)   # 15 / 6 = 2.5

hist(xbar_sim,
     breaks = 40,
     freq   = FALSE,
     main   = "Simulated vs. theoretical sampling distribution of x-bar",
     xlab   = "simulated sample mean")

curve(dnorm(x, mean = mu_true, sd = se_theoretical),
      add = TRUE,
      lwd = 2)
```

**This chunk is shown but not executed on this page** (`eval: false`) — running it yourself is the point. The
figure below was produced by running exactly this code separately (same seed, same `xbar_sim` as Step 3, same
theoretical curve), so you can check whether the curve sits on your own histogram the way it sits on this one:

![**Simulated histogram with the theoretical curve overlaid (synthetic).** The same 5000 simulated means as
the figure above, now with the theoretical Normal(48, 2.5²) density drawn on top. The curve sits closely
against the bars across the whole
range.](../assets/figures/lab02_sim_theoretical_overlay.png){#fig-lab02-theoretical-overlay fig-alt="The same bell-shaped histogram of simulated sample means as the previous figure, now with a smooth theoretical curve traced closely over the bars, peaking near 48."}

If the simulation and the theory agree, the smooth curve should sit almost exactly on top of the histogram's
bars, rising to a peak near x = 48 and falling off to near zero by roughly x = 40.5 and x = 55.5 (about three
theoretical standard errors, 3 × 2.5 = 7.5, on either side of 48).

## Verify

Before you consider this lab complete, check all of the following against your own output:

- `length(xbar_sim)` equals 5000 — one simulated mean per simulated sample, no more and no fewer.
- `mean(xbar_sim)` is close to 48 (the hypothetical true μ), not close to 49.8 (the real sample's x̄) — this
  simulation never touches the real MAC survey data, only the stipulated hypothetical world.
- `sd(xbar_sim)` is close to 2.5, matching σ/√n = 15/6 from the Week 2 derivation, not close to 15 (that would
  be the SD of a single *observation*, not of the sample *mean*).
- The histogram from Step 3 looks unimodal and roughly symmetric, centered near 48, before you have added any
  theoretical curve — the bell shape should emerge from the simulation alone.
- The overlaid curve from Step 4 sits close to the histogram's bars across the whole range, not systematically
  above or below them, and not shifted to one side.
- If you rerun the whole script from the very top with `set.seed(35103)` restored at each chunk, you get the
  same `xbar_sim` values and the same summary numbers every time.

As a second, unrelated context for the same idea, imagine (synthetic; seed 35103) a hypothetical true world of
package delivery times, Normal(μ = 120, σ = 24) minutes, and a sample size of n = 16 deliveries. The same
four-step recipe above — simulate B samples of size n, take `colMeans()`, summarize, overlay the theoretical
curve — would produce simulated means centered near 120 with an SD near σ/√n = 24/√16 = 24/4 = **6**. Nothing
about the four steps changes: only the three numbers you feed in (`mu_true`, `sigma_true`, `n`) change. That
portability — the same simulation recipe working for any Normal world and any sample size — is itself one of
the lab's lessons: a sampling-distribution simulation is a general tool, not something special to MAC visit
durations.

![**The same recipe, a different world (synthetic).** Running the identical four-step simulation with
`mu_true = 120`, `sigma_true = 24`, `n = 16` (same seed 35103, B = 5000) gives simulated means centered near
120 with SD near 6, matching the theoretical Normal(120, 6²) curve overlaid here — no part of the recipe
itself changed, only the three input
numbers.](../assets/figures/lab02_transfer_delivery.png){#fig-lab02-transfer-delivery fig-alt="A bell-shaped histogram of 5000 simulated sample means for a package-delivery-times example, centered near 120 minutes, with a smooth theoretical curve traced over it and summary text reporting the simulated mean and standard deviation."}

## AI use note

| Tool | Purpose | Verification |
|------|---------|--------------|
| AI coding assistant (e.g. Claude, ChatGPT, Copilot) | Drafting or troubleshooting base-R simulation code such as the `rnorm()` / `matrix()` / `colMeans()` pattern above, or explaining an unfamiliar base-R function's arguments | Run the AI-suggested code yourself in your own R session with `set.seed(35103)` set, and confirm `length(xbar_sim) == 5000`, `mean(xbar_sim)` near 48, and `sd(xbar_sim)` near 2.5 before trusting any AI-assisted rewrite; never submit AI output you have not run and checked against these numbers |

The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.

## See also

- [Week 2 — Sampling distributions and simulation](../notes/week-02-sampling-distributions-and-simulation.qmd)
  — the companion note deriving the Normal(48, 2.5²) sampling distribution this lab simulates.
- [Labs index](index.qmd) — the full list of course labs.
- [Notation glossary](../resources/notation-glossary.qmd) — definitions for μ, σ, x̄, n, and SE(x̄).
