---
title: "Lab 6 — Likelihood curves"
subtitle: "Plotting the Binomial likelihood by hand-in-code and reading off the maximum"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab is the computational companion to
> [Week 6 — Maximum likelihood estimation](../notes/week-06-maximum-likelihood-estimation.qmd). There, you
> derived the Binomial MLE by calculus: differentiate the log-likelihood with respect to $\pi$, set the
> derivative to zero, and solve to get $\hat\pi_{MLE} = k/n$. Here, you build the likelihood curve directly
> from the definition, over a grid of candidate $\pi$ values, and confirm **numerically** that the highest
> point on that curve sits exactly where the calculus said it would. Seeing the curve, and watching a search
> over candidate values land on the same answer as the derivative, is the point: calculus finds the peak of a
> curve, and this lab lets you look at the curve.

## The idea

A likelihood function $L(\pi)$ is not a probability distribution over $\pi$ — it is an ordinary function of
$\pi$, built from data that is already fixed. For the MAC usage-rate survey, the data in hand is fixed: $n =
100$ students surveyed, $k = 38$ of them used the MAC at least once this week. Once $n$ and $k$ are fixed, the
Binomial probability mass function

$$
P(K = k \mid \pi) = \binom{n}{k}\pi^k(1-\pi)^{n-k}
$$

can be re-read as a function of $\pi$ for that fixed $k$, and that re-reading is exactly $L(\pi)$. Because
$\binom{n}{k}$ does not involve $\pi$, it does not shift where the maximum falls, but this lab keeps it in
place anyway, so the curve you plot is the actual likelihood, not just the kernel.

The idea of this lab is almost embarrassingly direct: instead of taking a derivative, you can just **try a lot
of values of $\pi$, compute $L(\pi)$ (or, better, $\ell(\pi) = \ln L(\pi)$) at each one, and see which value
gives the largest result.** This is a grid search. It does not replace the calculus derivation from Week 6 —
it cannot tell you the maximum is a maximum in any exact sense, and it cannot hand you a clean closed-form
answer like $\hat\pi_{MLE} = k/n$. But it is an excellent way to *see* what the calculus derivation is doing,
and a useful sanity check whenever you are not sure you trust an algebraic derivative.

Two housekeeping points before you start.

First, **why work with $\ell(\pi) = \ln L(\pi)$ instead of $L(\pi)$ itself.** With $n = 100$, a term like
$\pi^{38}(1-\pi)^{62}$ is an extremely small number — computing it directly risks **numerical underflow**,
where a value that is mathematically positive gets rounded to exactly `0` in floating-point arithmetic because
it is too small for the computer's number representation to distinguish from zero. Logarithms turn that tiny
product into a sum of manageable-sized terms ($k\ln\pi + (n-k)\ln(1-\pi)$, plus a constant from the binomial
coefficient), which is numerically well behaved at any sample size. Because $\ln(\cdot)$ is a strictly
increasing function, the location of the maximum is identical whether you maximize $L(\pi)$ or $\ell(\pi)$ —
only the vertical scale changes. This lab plots $\ell(\pi)$ for exactly that numerical-stability reason, and
says so at the point it matters.

![**Why the log scale (synthetic).** The same MAC-survey likelihood, computed two ways across the full grid.
Left: the raw likelihood $L(\pi)$ spans roughly 150+ orders of magnitude (from about $8\times10^{-2}$ at the
peak down to about $5\times10^{-159}$ near $\pi=0.999$) — so small everywhere except right at the peak that an
ordinary plot shows it as flat at zero almost everywhere. Right: the log-likelihood $\ell(\pi)$ has the exact
same shape and the exact same peak location, but on a numerically well-behaved vertical scale (about $-2.5$ at
the peak down to about $-364$ at the far
edge).](../assets/figures/lab06_underflow_log_scale.png){#fig-lab06-underflow fig-alt="Two side-by-side line plots over pi from 0 to 1. The left panel, raw likelihood, shows a sharp narrow spike near pi equals 0.38 and looks essentially flat at zero everywhere else. The right panel, log-likelihood, shows the same single-peaked hill shape but fully visible across the whole range, from about negative 2.5 at the peak down to roughly negative 200 to negative 364 at the two edges."}

Second, a grid search finds the best value *among the grid points you tried* — it is only as fine as the grid.
A grid stepping by $0.001$ across $(0, 1)$ gives 999 candidate values, which is plenty of resolution to land on
$\pi = 0.380$ to three decimal places; a coarser grid (say, steps of $0.05$) would only be able to land on
$0.35$ or $0.40$, missing $0.38$ entirely. The finer the grid, the closer the numerical maximum sits to the
true calculus answer — in the limit of an infinitely fine grid, they coincide exactly.

![**Two paths to the same peak (synthetic).** The calculus route from Week 6 (set the derivative to zero and
solve) and the grid-search route this lab builds (try many candidate $\pi$ values and take the argmax) start
from the same fixed data and land on the same
answer.](../assets/figures/lab06_two_paths_pipeline.png){#fig-lab06-two-paths fig-alt="A branching flow diagram. A top box reads fixed data n equals 100, k equals 38. Two arrows lead down to two paths: a left path, calculus, set the derivative of the log-likelihood to zero and solve, ending in pi-hat equals k over n equals 0.38; and a right path, grid search, try about 1000 candidate pi values and take the argmax, ending in pi-hat equals 0.380 to grid resolution. A caption below reads: different methods, same question, the same answer."}

## Goal

Using the full MAC usage-rate survey ($n = 100$, $k = 38$), build a grid of candidate $\pi$ values, compute the
(log-)likelihood at each one, plot the resulting curve, and confirm that its numerical maximum sits at
$\hat\pi = 0.38$ — matching $k/n = 38/100 = 0.38$, the closed-form MLE derived by calculus in Week 6.

## Setup

- Base R only; no packages beyond what ships with R.
- `set.seed(35103)` is included in the setup chunk for consistency with the rest of the course's convention,
  even though this particular lab does not simulate random data — every number here (the grid, $n$, $k$) is
  fixed and deterministic, and the plot you produce will be identical every time you run it.
- You will need: `dbinom()` (density/mass function, here used with `log = TRUE`), `seq()` (to build the grid),
  `which.max()` (to locate the peak), and base graphics (`plot()`, `abline()`, `points()`).
- The two numbers that drive this whole lab are fixed by the MAC Study and must not be changed: $n = 100$
  survey respondents, $k = 38$ of whom used the MAC at least once this week.

```{r}
#| label: setup
#| eval: false
set.seed(35103)

n <- 100   # full MAC usage-rate survey: students surveyed
k <- 38    # students who used the MAC at least once this week
```

## Steps

### Step 1 — Define a grid of candidate $\pi$ values

A proportion can only take values in $[0, 1]$, so that is the natural range for the grid. Avoid the exact
endpoints $0$ and $1$: at $\pi = 0$ the term $\ln(\pi)$ is undefined (and $k > 0$, so $\pi = 0$ cannot be the
right answer anyway), and at $\pi = 1$ the term $\ln(1-\pi)$ is undefined (and $k < n$, so $\pi = 1$ cannot be
right either). A step of $0.001$ gives 999 candidate values between (but not including) $0$ and $1$, which is
fine enough resolution to recover $\hat\pi = 0.380$ to three decimal places.

```{r}
#| label: step1-grid
#| eval: false
pi_grid <- seq(0.001, 0.999, by = 0.001)

# sanity checks on the grid before using it
length(pi_grid)     # 999 candidate values
range(pi_grid)       # (0.001, 0.999) -- neither endpoint is exactly 0 or 1
```

### Step 2 — Compute the (log-)likelihood at each grid value

For each candidate $\pi$ in the grid, compute the log-likelihood of observing $k = 38$ successes in $n = 100$
trials, treating that candidate $\pi$ as if it were the true success probability. `dbinom(k, size = n, prob =
pi, log = TRUE)` does exactly this: it returns $\ln P(K = k \mid \pi) = \ln\!\binom{n}{k} + k\ln\pi +
(n-k)\ln(1-\pi)$, evaluated at whatever `pi` you hand it, for the fixed `k` and `n`. Because `dbinom()` is
vectorized, handing it the whole grid at once returns a same-length vector of log-likelihood values — one
per candidate $\pi$ — without writing an explicit loop.

```{r}
#| label: step2-loglik
#| eval: false
log_lik <- dbinom(x = k, size = n, prob = pi_grid, log = TRUE)

# a peek at the shape: log-likelihood is large-negative at the extremes...
head(data.frame(pi = pi_grid, log_lik = log_lik), 3)
tail(data.frame(pi = pi_grid, log_lik = log_lik), 3)

# ...and (we expect) least negative -- i.e. largest -- somewhere in the middle
summary(log_lik)
```

Two things are worth noticing in that output before moving on. Every value of `log_lik` is negative, because
`dbinom()` returns a log-*probability*, and a probability strictly between 0 and 1 always has a negative
natural log. And the values near the two ends of the grid (`pi` close to `0.001` or `0.999`) are large in
magnitude and very negative — those candidate values are extremely poor explanations of seeing 38 successes
out of 100 trials, and the log-likelihood says so emphatically.

### Step 3 — Plot the log-likelihood curve and mark its numerical maximum

`which.max()` returns the *position* in a vector where the largest value occurs — here, the grid index of the
candidate $\pi$ with the highest log-likelihood. Use that position to pull out both the best $\pi$ and its
log-likelihood, then draw the curve with that point marked.

```{r}
#| label: step3-plot
#| eval: false
best_index   <- which.max(log_lik)
pi_hat_grid  <- pi_grid[best_index]
loglik_at_hat <- log_lik[best_index]

pi_hat_grid     # the numerical MLE read off the grid
loglik_at_hat   # the log-likelihood value at that point

plot(pi_grid, log_lik, type = "l", lwd = 2,
     xlab = expression(pi), ylab = expression(paste("log-likelihood  ", ell(pi))),
     main = "Binomial log-likelihood, MAC usage survey (n = 100, k = 38)")
abline(v = pi_hat_grid, lty = 2, col = "gray40")
points(pi_hat_grid, loglik_at_hat, pch = 19, col = "firebrick", cex = 1.3)
```

**This chunk is shown but not executed on this page** (`eval: false`) — running it yourself is the point. The
figure below was produced by running exactly this grid search separately (same `pi_grid`, same `dbinom()`
call, same `which.max()` logic), so you can compare its shape and its marked peak to what you get when you run
it yourself:

![**The Binomial log-likelihood curve, MAC usage survey (synthetic).** A single smooth hill over $\pi \in
(0,1)$, rising from a very negative value near $\pi=0.001$, peaking at $\hat\pi=0.380$ (marked point), and
falling away toward $\pi=0.999$ — exactly the shape and peak location the calculus derivation
predicts.](../assets/figures/lab06_mac_loglik_curve.png){#fig-lab06-mac-curve fig-alt="A single-peaked line plot of log-likelihood against pi from 0 to 1, rising from a very negative value near pi equals 0, peaking near pi equals 0.38 where a red dot and a dashed vertical line mark the maximum, then falling away toward pi equals 1."}

The curve should look like a single smooth hill: rising from a very negative value near $\pi = 0.001$, reaching
one clear peak, and falling away again toward $\pi = 0.999$. That single-peaked (unimodal) shape is a general
feature of the Binomial log-likelihood, not a coincidence of these particular numbers — it is exactly what
lets a derivative-based search (Week 6's calculus) find the same maximum a grid search finds here.

### Step 4 — Confirm the numerical maximum against the calculus MLE

Week 6 derived, by setting $\dfrac{d\ell}{d\pi} = 0$ and solving, that the Binomial MLE has the closed form
$\hat\pi_{MLE} = k/n$. For this survey, $k/n = 38/100 = 0.38$. Compare that closed-form value directly against
`pi_hat_grid` from Step 3.

```{r}
#| label: step4-confirm
#| eval: false
pi_hat_calculus <- k / n
pi_hat_calculus     # 0.38, from Week 6's derivative-based derivation

pi_hat_grid         # 0.38 (to the grid's 0.001 resolution), from the search above

all.equal(pi_hat_grid, pi_hat_calculus, tolerance = 0.001)
```

With a grid step of $0.001$, `pi_hat_grid` lands at exactly `0.38`, matching `pi_hat_calculus` to the
grid's resolution. This is the payoff of the whole lab: two completely different methods — an algebraic
derivative set to zero, and a brute-force search over nearly a thousand candidate values — arrive at the same
answer, because both are answering the same question ("which $\pi$ makes the observed data most likely?").

## Verify

Work through this checklist before considering the lab complete.

- `length(pi_grid)` returns `999`, and `range(pi_grid)` shows neither endpoint is exactly `0` or `1`.
- `log_lik` is the same length as `pi_grid` (`length(log_lik) == length(pi_grid)` is `TRUE`), and every entry
  is negative.
- The plotted curve is single-peaked: it rises, reaches one maximum, and falls — with no second bump.
- `pi_hat_grid` equals `0.38` (to three decimal places), matching `k / n = 38 / 100 = 0.38` from Week 6.
- Nudging the grid finer (e.g. `by = 0.0001` instead of `by = 0.001`) still lands the maximum at `0.38` — the
  answer is not an artifact of a particular grid spacing.

### A transfer example — same idea, a different, synthetic context

To confirm you understand the *method* and not just this one dataset, repeat Steps 1–4 for a second,
unrelated binomial scenario: **synthetic** data (not the MAC Study) from a small-business quality-control
setting. A parts supplier tests a batch of $n = 50$ synthetic component samples for a defect, and finds $k =
6$ defective units. Under the same Binomial model, the calculus MLE is $\hat\pi_{MLE} = k/n = 6/50 = 0.12$.

```{r}
#| label: transfer-example
#| eval: false
set.seed(35103)

# synthetic quality-control data -- NOT the MAC Study, illustrative only
n_qc <- 50
k_qc <- 6

pi_grid_qc <- seq(0.001, 0.999, by = 0.001)
log_lik_qc <- dbinom(x = k_qc, size = n_qc, prob = pi_grid_qc, log = TRUE)

best_index_qc  <- which.max(log_lik_qc)
pi_hat_grid_qc <- pi_grid_qc[best_index_qc]

pi_hat_grid_qc          # should land at 0.12
k_qc / n_qc             # 6 / 50 = 0.12, the calculus MLE for comparison

plot(pi_grid_qc, log_lik_qc, type = "l", lwd = 2,
     xlab = expression(pi), ylab = expression(paste("log-likelihood  ", ell(pi))),
     main = "Synthetic transfer example: quality-control defect rate (n = 50, k = 6)")
abline(v = pi_hat_grid_qc, lty = 2, col = "gray40")
points(pi_hat_grid_qc, log_lik_qc[best_index_qc], pch = 19, col = "firebrick", cex = 1.3)
```

**This chunk is shown but not executed on this page** (`eval: false`) — running it yourself is the point. The
figure below was produced by running exactly this grid search separately (same synthetic `n_qc`/`k_qc`, same
`pi_grid_qc`, same `dbinom()`/`which.max()` logic), so you can compare it to your own run:

![**The Binomial log-likelihood curve, synthetic QC transfer example (synthetic).** The same single-peaked
shape as the MAC-survey curve, now for $n=50$, $k=6$, peaking at $\hat\pi=0.120$ (marked
point).](../assets/figures/lab06_qc_transfer_loglik_curve.png){#fig-lab06-qc-curve fig-alt="A single-peaked line plot of log-likelihood against pi from 0 to 1 for the quality-control transfer example, peaking near pi equals 0.12 where a red dot and a dashed vertical line mark the maximum."}

If `pi_hat_grid_qc` comes out to `0.12`, matching `k_qc / n_qc`, you have confirmed the grid-search method
generalizes beyond the MAC Study's particular numbers — it is a property of the Binomial likelihood in
general, not a coincidence of $n = 100, k = 38$.

![**The two worked curves, side by side (synthetic).** The MAC-survey curve ($n=100$, $\hat\pi=0.38$, width at
2 log-likelihood units below the peak $\approx 0.191$) and the QC-transfer curve ($n=50$, $\hat\pi=0.12$, same
width measure $\approx 0.181$), overlaid on shared axes. The QC curve is not the wider of the two, even though
it comes from fewer trials — its $\hat\pi$ sits much closer to the 0/1 boundary, which raises the
per-observation information enough to roughly offset its smaller $n$. Curve width is driven by both sample
size and where $\hat\pi$ falls, not by sample size
alone.](../assets/figures/lab06_curve_width_comparison.png){#fig-lab06-curve-comparison fig-alt="Two overlaid single-peaked line plots of log-likelihood against pi: a blue curve peaking near pi equals 0.38 for the MAC survey, and a rust-colored curve peaking near pi equals 0.12 for the QC transfer example, with dashed vertical lines at each peak and a legend reporting each curve's measured width near its peak."}

## AI use note

| Tool | Purpose | Verification |
|------|---------|---------------|
| AI coding assistant (e.g. an LLM-based chat or IDE completion tool) | Drafting or explaining a `dbinom()`/`which.max()` grid-search snippet, or debugging a plotting call | Re-derive the calculus MLE ($k/n$) independently and check the grid-search output against it by hand before trusting any AI-suggested code; re-run the chunk yourself line by line rather than pasting a block unread |

## See also

- [Week 6 — Maximum likelihood estimation](../notes/week-06-maximum-likelihood-estimation.qmd) — the calculus
  derivation this lab confirms numerically.
- [Week 5 — Likelihood](../notes/week-05-likelihood.qmd) — where the likelihood function itself was introduced,
  using the small five-response pilot batch.
- [Inference formula reference](../resources/inference-formula-reference.qmd)
- [Notation glossary](../resources/notation-glossary.qmd)
- [R and Quarto setup](../resources/r-quarto-setup.qmd)

The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.
